Alone Together: Compositional Reasoning and Inference for Weak Isolation

### 2017-10-22 implementation.tex
Stability definition changed.

\textbf{Real-World Isolation Levels} The axiomatization of the
isolation levels presented in \S\ref{sec:isolation} leave out
certain nuances of their implementations on real data stores, which
need to be taken into account for verification to be effective in
practice.  We consider these into account while linking \tool with
store-specific semantics (isolation specifications etc). As an
example, consider how PostgreSQL implements an \C{UPDATE} operation.
\C{UPDATE} first selects the records that meet the search criteria
from the snapshot against which it is executing (the snapshot is
established at the beginning of the transaction if the isolation level
is SI, or at the beginning of the \C{UPDATE} statement if the
isolation level is RC). The selected records are then visited in the
actual database (if they still exist), write locks are obtained, and
the update is performed, provided that the record still meets
\C{UPDATE}'s search criteria. If the record no longer meets the
search criteria (due to a concurrent update) the record is excluded
from the update, and the write lock is immediately released.
Otherwise, the record remains locked until the transaction commits.

>>>>

\textbf{Real-World Isolation Levels} The axiomatization of the
isolation levels presented in \S\ref{sec:isolation} leaves out
certain nuances of their implementations on real data stores, which
need to be taken into account for verification to be effective in
practice.  We consider these into account while linking \tool with
store-specific semantics (isolation specifications etc). As an
example, consider how PostgreSQL implements an \C{UPDATE} operation.
\C{UPDATE} first selects the records that meet the search criteria
from the snapshot against which it is executing (the snapshot is
established at the beginning of the transaction if the isolation level
is SI, or at the beginning of the \C{UPDATE} statement if the
isolation level is RC). The selected records are then visited in the
actual database (if they still exist), write locks are obtained, and
the update is performed, provided that the record still meets
\C{UPDATE}'s search criteria. If the record no longer meets the
search criteria (due to a concurrent update) the record is excluded
from the update, and the write lock is immediately released.
Otherwise, the record remains locked until the transaction commits.

### 2017-10-22 motivation.tex
Stability definition changed.

As this discussion hints, implementations of SI on real databases
such as PostgresSQL are highly complicated, often running into
thousands of lines of code.  Nonetheless, the semantics of SI, in
terms of how it effects transitions on the database state, can be
captured in a fairly simple model. First, effects induced by one
transaction (call it $T$) are not visible to another concurrently executing one
during $T$'s execution. Thus, from $T$'s perspective,
the global state does not change during its execution.  More formally,
for every operation performed by $T$, the global state $T$ witnesses
before ($\stg$) and after ($\stg')$ executing the operation is the
same ($\stg'=\stg$).  After $T$ finishes execution, it commits its
changes to the actual database, which may have already incorporated
the effects of concurrent transactions. In executions where $T$
successfully commits, concurrent transactions are guaranteed to not be
in write-write conflict with $T$. Thus, if $\stg$ is the global state
that $T$ witnessed when it finished execution (the snapshot state),
and $\stg'$ is the state to which $T$ commits, then the difference
between $\stg$ and $\stg'$ should not result in a write-write conflict
with $T$. To concretize this notion, let the database state be a map
from database locations to values, and let $\stl$ denote a
transaction-local log that maps the locations being written to their
updated values.  The absence of write-write conflicts between $T$ and
the diff between $\stg$ and $\stg'$ can be expressed as: $\forall
x\in\mathit{dom}(\delta)$, $\stg'(x) = \stg(x)$.  In other words, the
semantics of SI can be captured as an axiomatization over transitions
of the database state ($\Delta \longrightarrow \Delta'$) during a
transaction's ($T$) lifetime:
\begin{itemize}
  \item While $T$ executes, $\Delta' = \Delta$.
  \item After $T$ finishes execution, but before it commits its local
    state $\delta$, $\forall(x\in\mathit{dom}(\delta)).~\Delta'(x) = \Delta(x)$.
\end{itemize}

@@

\noindent This simple characterization of SI isolation allows us to
verify the consistency conditions associated with the \C{new\_order}
transaction.  First, since the database does not change ($\Delta' =
\Delta$) during execution of the transaction's body, we can reason
about \C{new\_order} as though it executed in complete isolation until
its commit point, leading to a verification process similar to what
would been applied when reasoning sequentially.  When \C{new\_order}
finishes execution, however, but before it commits, the SI
axiomatization shown above requires us to consider global state
transitions $\stg \longrightarrow \stg'$ that do not include changes
to the records ($\stl$) written by \C{new\_order}, i.e.,
$\forall(x\in\mathit{dom}(\delta)).~\Delta'(x) = \Delta(x)$.  The axiomatization
precludes any execution in which there are concurrent updates to
shared table fields (e.g., \C{d\_next\_o\_id} on the same \C{District}
table), but does not prohibit interferences that write to different
tables, or write to different records in the same table.  We need to
reason about the safety of such interferences with respect to
\C{new\_order}'s consistency invariants.

>>>>

As this discussion hints, implementations of SI on real databases
such as PostgreSQL are highly complicated, often running into
thousands of lines of code.  Nonetheless, the semantics of SI, in
terms of how it effects transitions on the database state, can be
captured in a fairly simple model. First, effects induced by one
transaction (call it $T$) are not visible to another concurrently executing one
during $T$'s execution. Thus, from $T$'s perspective,
the global state does not change during its execution.  More formally,
for every operation performed by $T$, the global state $T$ witnesses
before ($\stg$) and after ($\stg')$ executing the operation is the
same ($\stg'=\stg$).  After $T$ finishes execution, it commits its
changes to the actual database, which may have already incorporated
the effects of concurrent transactions. In executions where $T$
successfully commits, concurrent transactions are guaranteed to not be
in write-write conflict with $T$. Thus, if $\stg$ is the global state
that $T$ witnessed when it finished execution (the snapshot state),
and $\stg'$ is the state to which $T$ commits, then the difference
between $\stg$ and $\stg'$ should not result in a write-write conflict
with $T$. To concretize this notion, let the database state be a map
from database locations to values, and let $\stl$ denote a
transaction-local log that maps the locations being written to their
updated values.  The absence of write-write conflicts between $T$ and
the diff between $\stg$ and $\stg'$ can be expressed as: $\forall
x\in\mathit{dom}(\delta)$, $\stg'(x) = \stg(x)$.  In other words, the
semantics of SI can be captured as an axiomatization over transitions
of the database state ($\Delta \longrightarrow \Delta'$) during a
transaction's ($T$) lifetime:
\begin{itemize}
  \item While $T$ executes, $\Delta' = \Delta$.
  \item After $T$ finishes execution, but before it commits its local
    state $\delta$, $\forall(x\in\mathit{dom}(\delta)).~\Delta'(x) = \Delta(x)$.
\end{itemize}

@@

\noindent This simple characterization of SI isolation allows us to
verify the consistency conditions associated with the \C{new\_order}
transaction.  First, since the database does not change ($\Delta' =
\Delta$) during execution of the transaction's body, we can reason
about \C{new\_order} as though it executed in complete isolation until
its commit point, leading to a verification process similar to what
would have been applied when reasoning sequentially.  When
\C{new\_order} finishes execution, however, but before it commits, the
SI axiomatization shown above requires us to consider global state
transitions $\stg \longrightarrow \stg'$ that do not include changes
to the records ($\stl$) written by \C{new\_order}, i.e.,
$\forall(x\in\mathit{dom}(\delta)).~\Delta'(x) = \Delta(x)$.  The
axiomatization precludes any execution in which there are concurrent
updates to shared table fields (e.g., \C{d\_next\_o\_id} on the same
\C{District} table), but does not prohibit interferences that write to
different tables, or write to different records in the same table.  We
need to reason about the safety of such interferences with respect to
\C{new\_order}'s consistency invariants.

### 2017-10-22 reasoning.tex
Stability definition changed.

A central feature of a rely-guarantee judgment is a stability
condition that requires the validity of an assertion $\phi$ to be
unaffected by interference from other concurrently executing
transactions, i.e., the rely relation $R$. In conventional RG,
stability is defined as follows, where $\sigma$ and $\sigma'$ denote
states:
\begin{smathpar}
\begin{array}{lcl}
\stable(R,\phi) & \Leftrightarrow & \forall \sigma,\sigma'.~
\phi(\sigma) \conj R(\sigma,\sigma') \Rightarrow \phi(\sigma')\\
\end{array}
\end{smathpar}
Due to the presence of local and global database states, and the
availability of an isolation specification, we use multiple
definitions of stability in Fig.~\ref{fig:rg-rules}, but they all
convey the same intuition as above. In our setting, we only need to
prove the stability of an assertion ($\phi$) against those environment
steps which lead to a global database state on which the transaction
itself can take its next step according to its isolation specification
($\I$). 
\begin{smathpar}
\begin{array}{lcl}
\stable(R, \phi) & \Leftrightarrow & \forall \stl, \stg, \stg'. \phi(\stl, \stg) \wedge R^{*}(\stg, \stg') \wedge \I(\stl, \stg, \stg') \Rightarrow \phi(\stl, \stg')
\end{array}
\end{smathpar}
\noindent A characteristic of RG reasoning is that stability of an assertion is
always proven w.r.t to $R$, and not $R^{*}$, although interference may
include multiple environment steps, and $R$ only captures a single
step. This is nonetheless sound due to inductive reasoning: if $\phi$
is preserved by every step of $R$, then $\phi$ is preserved by
$R^{*}$, and vice-versa.  However, such reasoning does not extend
naturally to isolation-constrained interference because $R^{*}$ modulo
$\I$ is not same as $\R^{*}$; the former is a transitive relation
constrained by $\I$, whereas the latter is the transitive closure of
a constrained relation. This means, unfortunately, that we cannot directly replace $R^{*}$
by $R$ in the above condition because of our need to consider the weak
isolation specification $\I$ that selectively exposes interferences.

To obtain an equivalent form in our setting, we require an additional
condition on the isolation specification, which we call the
\emph{stability condition on $\I$}.  The condition requires $\I$ to
allow interferences $R^{*}(\stg,\stg'')$, for two database states
$\stg$ and $\stg''$, only if it also allows interference for every
prefix of $R^{*}(\stg,\stg'')$. In other words, if $\I$ disallows
interference from $\stg$ to $\stg'$, then an $R$-step from $\stg'$ to
$\stg''$ should not make the interference from $\stg$ to $\stg''$
valid.  This stability condition on $\I$ is defined formally thus:
\begin{smathpar}
\begin{array}{lcl}
  \stable(R,\I) & \Leftrightarrow & \forall \stl,\stg,\stg',\stg''.~
  \neg\I(\stl,\stg,\stg') \conj R(\stg',\stg'') \Rightarrow
  \neg\I(\stl,\stg,\stg'')
\end{array}
\end{smathpar}
It can be easily verified that the above stability condition is
satisfied by the isolation axioms from Sec.~\ref{sec:isolation}. For
instance, $\I_{ss}$, the snapshot axiom, is stable because if
$\I_{ss}$ is invalid ($\neg\I_{ss}$), then an interference has already
modified a record, and no further interference will restore the
original record, because new commits bear new transaction ids
different from the transaction id (\C{txn} field) of the original
record. We can now generalize the stability condition for any
arbitrary assertion $\phi$ by only considering those environment steps
which in addition to preserving $\phi$, also preserve the condition
that the transaction can continue to take its next step, i.e. the
condition $\I$ holds on both the beginning and the end states.
 \begin{smathpar}
   \forall \stl, \stg, \stg', \stg_1. \phi(\stl, \stg) \wedge \I(\stl, \stg_1, \stg) \wedge R^{*}(\stg, \stg') \wedge \I(\stl, \stg_1, \stg')
   \Rightarrow \phi(\stl, \stg')
\end{smathpar}
Here, we use a separate notation ($\stg_1$) for the global database
state that the transaction witnessed during its last step. $\phi$ does
not constrain this global database state, but the stability of $\I$
ensures that any database state constrained by $\phi$ must also obey
$\I$. We can now replace the transitive closure in the rule with the
single step relation $R$, backed by the inductive argument for
stability. Pushing the universal quantifier for $\stg_1$ inside the
antecedent and separating the predicates involving $\stg_1$, we can
rewrite the condition in the form below, by defining a new ternary
relation $R \backslash \I$, which can be thought of as $R$ modulo
$\I$:
\begin{smathpar}
\begin{array}{lcl}
\stable((R \backslash \I), \phi) & \Leftrightarrow & \forall \stl,\stg,\stg'. \phi(\stl,\stg) \wedge (R \backslash \I)(\stl,\stg,\stg') \Rightarrow \phi(\stl,\stg')\\
(R \backslash \I)(\stl, \stg, \stg') & \Leftrightarrow & \exists \stg_1. R(\stg, \stg') \wedge \I(\stl, \stg_1, \stg) \wedge \I(\stl, \stg_1, \stg')
\end{array}
\end{smathpar}
This condition often significantly simplifies the form of $R
\backslash \I$ irrespective of $R$. For example, when a transaction is
executed against a snapshot of the database (i.e. $\mathbb{I}_{ss}$),
$R \backslash \I_{ss}$ will be the identity function, since if $\stg
\neq \stg'$, there does not exist any $\stg_1$ which satisfies the
constraint for $(R \backslash \I_{ss})(\stl,\stg,\stg')$.

@@

\paragraph{{\sc Proof Sketch.}}  The most important rule is the top-level
rule \rulelabel{RG-Txn}, which proves that a transaction $c$ which
begins its execution in global database state satisfying $I$ and
encountering interference $R$ while executing under isolation
specification $\I$ finishes its execution in a database state also
satisfying $I$, and also guarantees that its commit step satisfies
$G$. The rule uses the transaction-local RG judgement $\R_e \vdash
\rg{P}{c}{Q}$. By \rulelabel{E-Txn-Start}, the local and global
database states at the start of a transaction satisfy $P$, and the
only challenge is that environment steps in an execution covered by
$\R_e \vdash \rg{P}{c}{Q}$ are in $\R_e$, while the top-level
judgement requires environment steps in $R$. We show that it is enough
to consider only those enviroment steps in $\R_e$. First, we use an
inductive argument and stability of $\I_e$ ($\stable(R, \I_e)$) to
show that any execution in which the transaction completes all its
steps must always preserve the isolation specification $\I_e$ after
every environment step. Intuitively, this is because once $\I_e$ gets
broken after some environment step, it will continue to remain broken
and the transaction would not be able to proceed (according to
\rulelabel{E-Txn}). Since $\R_e$ contains exactly those environment
steps which preserve $\I_e$, the local-level RG judgement can be
soundly used, which guarantees that after the transaction finishes its
execution, its local state $\stl$ and global state $\stg$ will satisfy
the assertion $Q$. Environment steps between the last step of the
transaction and its commit step can modify the global state, and hence
we also require $Q$ to be stable againt $R$. Again, we use an
inductive argument, the stability of $\I_c$, and the fact that the
transaction must execute its commit step to show that all environment
steps must preserve $\I_c$, and hence it is enough to require
$\stable(\R_c, Q)$. $Q$ guarantees that the commit step is in $G$, and
$G$ in turn guarantees that after execution, the global database state
will obey the invariant $I$.

>>>>

A central feature of a rely-guarantee judgment is a stability
condition that requires the validity of an assertion $\phi$ to be
unaffected by interference from other concurrently executing
transactions, i.e., the rely relation $R$. In conventional RG,
stability is defined as follows, where $\sigma$ and $\sigma'$ denote
states:
\begin{smathpar}
\begin{array}{lcl}
\stable(R,\phi) & \Leftrightarrow & \forall \sigma,\sigma'.~
\phi(\sigma) \conj R(\sigma,\sigma') \Rightarrow \phi(\sigma')\\
\end{array}
\end{smathpar}
Due to the presence of local and global database states, and the
availability of an isolation specification, we use multiple
definitions of stability in Fig.~\ref{fig:rg-rules}, but they all
convey the same intuition as above. In our setting, we only need to
prove the stability of an assertion ($\phi$) against those environment
steps which lead to a global database state on which the transaction
itself can take its next step according to its isolation specification
($\I$). 
\begin{smathpar}
\begin{array}{lcl}
\stable(R, \phi) & \Leftrightarrow & \forall \stl, \stg, \stg'. \phi(\stl, \stg) \wedge R^{*}(\stg, \stg') \wedge \I(\stl, \stg, \stg') \Rightarrow \phi(\stl, \stg')
\end{array}
\end{smathpar}
\noindent A characteristic of RG reasoning is that stability of an
assertion is always proven w.r.t to $R$, and not $R^{*}$, although
interference may include multiple environment steps, and $R$ only
captures a single step. This is nonetheless sound due to inductive
reasoning: if $\phi$ is preserved by every step of $R$, then $\phi$ is
preserved by $R^{*}$, and vice-versa.  However, such reasoning does
not extend naturally to isolation-constrained interference because
$R^{*}$ modulo $\I$ is not same as $\R^{*}$; the former is a
transitive relation constrained by $\I$, whereas the latter is the
transitive closure of a relation constrained by $\I$. This means,
unfortunately, that we cannot directly replace $R^{*}$ by $R$ in the
above condition because of our need to consider the weak isolation
specification $\I$ that selectively exposes interferences.

To obtain an equivalent form in our setting, we require an additional
condition on the isolation specification, which we call the
\emph{stability condition on $\I$}.  The condition requires $\I$ to
admit the interference of multiple $R$ steps (i.e.,
$R^{*}(\stg,\stg'')$, for two database states $\stg$ and $\stg''$),
only if it also admits interference of each $R$ step along the way.
Formally:
\begin{smathpar}
\begin{array}{lcl}
  \stable(R,\I) & \Leftrightarrow & \forall \stl,\stg,\stg',\stg''.~
  \I(\stl,\stg,\stg'') \conj R(\stg',\stg'') \Rightarrow
  \I(\stl,\stg,\stg') \conj \I(\stl,\stg',\stg'')
\end{array}
\end{smathpar}
It can be easily verified that the above stability condition is
satisfied by the isolation axioms from Sec.~\ref{sec:isolation}. For
instance, $\I_{ss}$, the snapshot axiom, is stable because if a the
state is unmodified between $\stg$ and $\stg''$, then it is clearly
unmodified between $\stg$ and $\stg'$, and also between $\stg'$ and
$\stg''$, where $\stg'$ is an intermediary state.  Modifying and
restoring the state $\stg$ is not possible because each new commit
bears a new transaction id different from the transaction ids (\C{txn}
fields) present in $\stg$.

The stability condition on $\I$ guarantees that an interference from
$R^*$ is admissible only if the interference due to each individual
$R$ step is admissible. In other words, it makes isolation-constrained
$R^*$ relation equal to the transitive closure of the
isolation-constrained $R$ relation. We call $R$ constrained by
isolation $\I$ as $R$ modulo $\I$ ($R\backslash \I$; written shortly
as $\R$), which is the following ternary relation:
\begin{smathpar}
  \begin{array}{lcl}
    (R \backslash \I)(\stl, \stg, \stg') & \Leftrightarrow & R(\stg,
      \stg') \wedge \I(\stl, \stg, \stg')\\
  \end{array}
\end{smathpar}
It is now enough to prove the stability of an RG assertion $\phi$
w.r.t $R\backslash \I$:
\begin{smathpar}
  \begin{array}{lcl}
    \stable((R \backslash \I), \phi) & \Leftrightarrow & \forall
      \stl,\stg,\stg'.~ \phi(\stl,\stg) \wedge (R \backslash \I)(\stl,\stg,\stg') \Rightarrow \phi(\stl,\stg')\\
  \end{array}
\end{smathpar}
This condition often significantly simplifies the form of $R
\backslash \I$ irrespective of $R$. For example, when a transaction is
executed against a snapshot of the database (i.e. $\mathbb{I}_{ss}$),
$R \backslash \I_{ss}$ will be the identity function, since any
non-trivial interference will violate the $\stg' = \stg$ condition
imposed by $\I_{ss}$.

@@

\paragraph{{\sc Proof Sketch.}}  The most important rule is the top-level
rule \rulelabel{RG-Txn}, which proves that a transaction $c$ which
begins its execution in global database state satisfying $I$ and
encountering interference $R$ while executing under isolation
specification $\I$ finishes its execution in a database state also
satisfying $I$, and also guarantees that its commit step satisfies
$G$. The rule uses the transaction-local RG judgment $\R_e \vdash
\rg{P}{c}{Q}$. By \rulelabel{E-Txn-Start}, the local and global
database states at the start of a transaction satisfy $P$, and the
only challenge is that environment steps in an execution covered by
$\R_e \vdash \rg{P}{c}{Q}$ are in $\R_e$, while the top-level
judgment requires environment steps in $R$. We show that it is enough
to consider only those environment steps in $\R_e$. First, we use an
inductive argument and stability of $\I_e$ ($\stable(R, \I_e)$) to
show that any execution in which the transaction completes all its
steps must always preserve the isolation specification $\I_e$ after
every environment step. Intuitively, this is because once $\I_e$ gets
broken after some environment step, it will continue to remain broken
and the transaction would not be able to proceed (according to
\rulelabel{E-Txn}). Since $\R_e$ contains exactly those environment
steps which preserve $\I_e$, the local-level RG judgment can be
soundly used, which guarantees that after the transaction finishes its
execution, its local state $\stl$ and global state $\stg$ will satisfy
the assertion $Q$. Environment steps between the last step of the
transaction and its commit step can modify the global state, and hence
we also require $Q$ to be stable against $R$. Again, we use an
inductive argument, the stability of $\I_c$, and the fact that the
transaction must execute its commit step to show that all environment
steps must preserve $\I_c$, and hence it is enough to require
$\stable(\R_c, Q)$. $Q$ guarantees that the commit step is in $G$, and
$G$ in turn guarantees that after execution, the global database state
will obey the invariant $I$.
